---
title : "R_assignment_final"
author: "SwathiNadendla"
date: "2025-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
```

# Part 1: Data Inspection

## Step 1: Read the Input Files
```{r}
genotypes <- read_tsv("fang_et_al_genotypes.txt", col_types = cols(.default = "c"))
positions <- read_tsv("snp_position.txt", col_types = cols(.default = "c"))

# CSV files for inspection for easier processing
write.csv(genotypes, "readable_fang_et_al_genotypes.csv", row.names = FALSE)
write.csv(positions, "readable_snp_position.csv", row.names = FALSE)
```

## Step 2: Extract Specific Columns from Positions
```{r}
cut_snp_position <- select(positions, SNP_ID, Chromosome, Position)
```

## Step 3: Separate Maize and Teosinte Data
```{r}
maize_groups <- c("ZMMIL", "ZMMLR", "ZMMMR")
teosinte_groups <- c("ZMPBA", "ZMPIL", "ZMPJA")

process_group_data <- function(genotypes, groups, prefix) {
  group_data <- genotypes %>% filter(Group %in% groups)
  transposed_data <- t(group_data)
  transposed_data <- as.data.frame(transposed_data, stringsAsFactors = FALSE)
  colnames(transposed_data) <- transposed_data[1, ]
  transposed_data <- transposed_data[-1, ]
  transposed_data <- cbind(SNP_ID = rownames(transposed_data), transposed_data)
  rownames(transposed_data) <- NULL
  joined_data <- inner_join(cut_snp_position, transposed_data, by = "SNP_ID")
  write.csv(joined_data, paste0("joined_", prefix, "_snp.csv"), row.names = FALSE)
  unknown_data <- joined_data %>% filter(grepl("unknown", Chromosome) | grepl("unknown", Position))
  write.csv(unknown_data, paste0("unknown_joined_", prefix, "_snp.csv"), row.names = FALSE)
  multiple_data <- joined_data %>% filter(grepl("multiple", Chromosome) | grepl("multiple", Position))
  write.csv(multiple_data, paste0("multiple_joined_", prefix, "_snp.csv"), row.names = FALSE)
  
  output_files <- lapply(1:10, function(chr) {
    chr_data <- joined_data %>% filter(Chromosome == as.character(chr))
    chr_data$Position <- suppressWarnings(as.numeric(chr_data$Position))
    chr_data <- chr_data[!is.na(chr_data$Position), ]
    
    if (nrow(chr_data) == 0) {
      return(NULL)
    }
    
    asc_data <- chr_data %>% arrange(Position)
    asc_file <- paste0(prefix, "_chr", chr, "_asc.csv")
    write.csv(asc_data, asc_file, row.names = FALSE)
    
    if (all(diff(asc_data$Position) >= 0)) {
      message(paste0("Ascending sort successful for chromosome ", chr))
    } else {
      message(paste0("Ascending sort failed for chromosome ", chr))
    }
    
    desc_data <- chr_data %>% arrange(desc(Position))
    desc_file <- paste0(prefix, "_chr", chr, "_desc.csv")
    write.csv(desc_data, desc_file, row.names = FALSE)
    
    if (all(diff(desc_data$Position) <= 0)) {
      message(paste0("Descending sort successful for chromosome ", chr))
    } else {
      message(paste0("Descending sort failed for chromosome ", chr))
    }
    
    desc_data[desc_data == "?/?"] <- "-/-"
    edited_desc_file <- paste0("edited_", prefix, "_chr", chr, "_desc.csv")
    write.csv(desc_data, edited_desc_file, row.names = FALSE)
    
    return(list(asc_file = asc_file, desc_file = desc_file, edited_desc_file = edited_desc_file))
  })
  return(output_files)
}

maize_output_files <- process_group_data(genotypes, maize_groups, "maize")
teosinte_output_files <- process_group_data(genotypes, teosinte_groups, "teosinte")

list(maize_output_files, teosinte_output_files)
```

# Part 2: Data Visualization

## Load Necessary Libraries
```{r}
library(ggplot2)
```

## Read the Input Files
```{r}
genotypes <- read_tsv("fang_et_al_genotypes.txt", col_types = cols(.default = "c"))
positions <- read_tsv("snp_position.txt", col_types = cols(.default = "c"))

positions <- positions %>%
  mutate(Position = as.numeric(Position)) %>%
  filter(!Chromosome %in% c("multiple", "unknown", "NA"),
         !is.na(Position))

genotypes_long <- genotypes %>%
  pivot_longer(cols = -c(Sample_ID, Group), names_to = "SNP_ID", values_to = "Genotype")

combined_data <- inner_join(genotypes_long, positions, by = "SNP_ID")

combined_data <- combined_data %>%
  mutate(Genotype_Type = case_when(
    Genotype %in% c("A/A", "C/C", "G/G", "T/T") ~ "Homozygous",
    Genotype == "-" ~ "Missing",
    TRUE ~ "Heterozygous"
  ))

maize_groups <- c("ZMMIL", "ZMMLR", "ZMMMR")
teosinte_groups <- c("ZMPBA", "ZMPIL", "ZMPJA")

maize_data <- combined_data %>% filter(Group %in% maize_groups)
teosinte_data <- combined_data %>% filter(Group %in% teosinte_groups)
```

## Plot Genotype Frequencies
```{r}
genotype_frequencies <- combined_data %>%
  group_by(Genotype_Type, Group) %>%
  summarize(Frequency = n()) %>%
  mutate(Proportion = Frequency / sum(Frequency))

ggplot(genotype_frequencies, aes(x = Group, y = Proportion, fill = Genotype_Type)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Distribution of Genotype Frequencies",
       x = "Group",
       y = "Proportion",
       fill = "Genotype Type") +
  scale_fill_manual(values = c("Homozygous" = "blue", "Heterozygous" = "green", "Missing" = "red"))
```

## Plot Boxplot of SNP Positions Across Chromosomes
```{r}
ggplot(combined_data, aes(x = Chromosome, y = Position, fill = Group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of SNP Positions Across Chromosomes",
       x = "Chromosome",
       y = "Position",
       fill = "Group") +
  scale_fill_manual(values = c("ZMMIL" = "blue", "ZMMLR" = "green", "ZMMMR" = "red",
                               "ZMPBA" = "purple", "ZMPIL" = "orange", "ZMPJA" = "pink"))
```

## Create Facetted Density Plot for SNP Positions on Each Chromosome
```{r}
ggplot(data = combined_data, aes(x = Position, fill = factor(Chromosome))) +
  geom_density(alpha = 0.5) +
  scale_fill_discrete(name = "Chromosome") +
  labs(x = "SNP Position", y = "Density") +
  facet_wrap(~Chromosome, ncol = 5) +
  theme_minimal()
```

## Calculate Summary Statistics for Each Sample
```{r}
sample_summary_data <- combined_data %>%
  group_by(Sample_ID) %>%
  summarize(
    total_sites = n(),
    homozygous_sites = sum(Genotype_Type == "Homozygous"),
    heterozygous_sites = sum(Genotype_Type == "Heterozygous"),
    missing_sites = sum(Genotype_Type == "Missing"),
    prop_homozygous = homozygous_sites / total_sites,
    prop_heterozygous = heterozygous_sites / total_sites,
    prop_missing = missing_sites / total_sites
  )
```

## Plot Proportions of Homozygous, Heterozygous, and Missing Data by Sample
```{r}
ggplot(sample_summary_data, aes(x = Sample_ID)) +
  geom_bar(aes(y = prop_homozygous, fill = "Homozygous"), alpha = 0.8, stat = "identity") +
  geom_bar(aes(y = prop_heterozygous, fill = "Heterozygous"), alpha = 0.8, stat = "identity") +
  geom_bar(aes(y = prop_missing, fill = "Missing"), alpha = 0.8, stat = "identity") +
  scale_fill_manual(values = c("Homozygous" = "darkgreen", "Heterozygous" = "yellow", "Missing" = "gray")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Sample ID", y = "Proportion", fill = "Site Type", 
       title = "Proportion of Homozygous, Heterozygous, and Missing Sites by Sample ID")
```